Ionization of biological molecules by multicharged ions by using the stoichiometric model

### 2019-09-20 NewMonti25.tex
added angle data + mol bind ener analysis

\title{Ionization of molecules by multicharged bared ions by using the
stoichiometric model}
\author{author}
\affil{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio (CONICET-UBA).
Casilla de correo 67, sucursal 28 (C1428EGA) Buenos Aires, Argentina.}
%\author{author}
%\affil{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio (CONICET-UBA).
%Casilla de correo 67, sucursal 28 (C1428EGA) Buenos Aires, Argentina.}
%\affil{Fac. de Ciencias Exactas y Naturales, Universidad de Buenos Aires}
\date{\today }

@@

The damage caused by the impact of multicharged heavy projectiles has 
become a field of interest due to its recent implementation in ion-beam 
cancer therapy. The effectiveness of the radiation depends on the 
choice of the ions. In particular, theoretical and experimental studies 
with different projectiles have concluded that charged carbon ions 
could be the most suitable ions to use. The study of such systems 
represent a challenge from the theoretical point of view. The most 
widely used method to compute ionization of multicharged atoms is the 
first Born approximation. 
This perturbative method warrants the $Z^{2}$ laws, where $Z$ is the 
projectile charge. However, the damage concentrated in the vicinities 
of the Bragg peak correspond to energies of hundreds of KeV/amu. 
Precisely in this region, often refered to as the intermediate energy 
region, the Born approximation starts to fail. Another theoretical 
issue arises from the targets themselves; we are dealing with complex 
molecules. This article deals precisely with these two aspects. 
First, we perform more appropriate calculations on the primary damage 
mechanism, i.e., atomic ionization by multicharged ions, which can 
replace the Born results. Second, we inspect and use an stoichiometric 
model, which reduces a molecule to a sum of atomic processes quantities 
weighted by the number of such atom in the molecule.

@@

@@

Lets us consider a molecule $M$ composed by $n_{\alpha}$ atoms of the
element $\alpha$, the SSM describes the total ionization cross section 
of the molecule $\sigma_{M}$ as a simple sum of ionization cross 
sections of the isolated atoms $\sigma_{\alpha}$, 
\begin{equation}
 \sigma_{M}=\sum\limits_{\alpha}n_{\alpha}\sigma_{\alpha}\,.  
 \label{eq:sumion}
\end{equation}
We divided sixteen molecular targets of our interest in three families. 
The classification defined is given in Table~\ref{tab:families}.

@@

The first attempt to develop a comprehensive but straightforward 
phenomenological model for electron ejection from large molecules was 
proposed by Toburen {\it et al.}~\cite{toburen1975,toburen1976}. 
The authors found it convenient to scale the experimental ionization 
cross section in terms of the number of active or weakly--bound valence 
electrons $n_e$ (i.e. total number of electrons minus the K--shell).
Following Toburen, we can define the ionization cross section per weakly 
bound electron, $\sigma_{e}^T$, as
\begin{equation}
\sigma_{e}^T=\frac{\sigma_{M}}{n_e}=\frac{\sum\limits_{\alpha}
n_{\alpha}\sigma_{\alpha}}{\sum\limits_{\alpha}n_{\alpha}\nu_{\alpha}^T}
=\sigma_{e}^T(v)\,, 
\label{27} 
\end{equation}
where $\nu_{\alpha}^T$ are the Toburen numbers for each $\alpha$ atom 
is given by
\begin{equation}
\nu_{\alpha}^T=\left\{ 
\begin{array}{ll}
1, & \text{for H,} \\
4, & \text{for C,} \\ 
5, & \text{for N and P,} \\ 
6, & \text{for O and S}\,.
\end{array}\right.
\label{eq:nelec} 
\end{equation} 
The Toburen rule can be stated by saying that 
$\sigma_{e}$ is a \textit{universal} parameter independent on the 
molecule, which depends solely on the impact velocity, and holds for 
high impact energies (i.e. 0.25--5 Mev/amu).
%In Ref.~\cite{toburen1976}, it was found experimentally that for proton 
%impact on some simple molecules, $\sigma_{e}$ results
%\begin{equation}
%\begin{array}{cccc}
%E\text{(MeV)}               & 0.25 & 1.00 & 2.00 \\ 
%\sigma_{e} (10^{-16}cm^{2}) & 0.39 & 0.17 & 0.11
%\end{array}
%\label{30}
%\end{equation}
%with an estimated error of about 20\%. A similar ratio was found in 
%Ref.~\cite{itoh2013} for proton impact on uracil and adenine. 
These $\nu_{\alpha}^T$ can be interpreted as the number of active 
electrons in the collision. Of course, at very high energies also de 
K--shell electrons will be ionized and these numbers will be different.
A similar dependence with the number of weakly bound electrons was 
found in Ref.~\cite{itoh2013} for proton impact on uracil and adenine.

@@

In a given biological medium, direct ionization by ion impact accounts 
for just a fraction of the overall damage. Secondary electrons, as well 
as recoil target ions, also contribute substantially to the total damage. 
We can consider the single differential cross section of the shell 
$nl$ of the atom $\alpha$, $d\sigma _{\alpha nl}/dE$, to be a function 
of the ejected electron energy $E$ as a simple distribution 
function~\cite{surdutovic2018}. Then, we can define the mean value 
$\overline{E}_{\alpha}$ as in Ref.~\cite{abril2015},
\begin{eqnarray}
\overline{E}_{\alpha} &=&\frac{\langle E_{\alpha}\rangle}{\langle
1\rangle}=\frac{1}{\sigma_{\alpha}}\sum\limits_{nl}\int dE\,E
\frac{d\sigma_{\alpha,nl}}{dE}\,,  
\label{40} \\
\langle 1\rangle &=&\sigma_{\alpha}=\sum\limits_{nl}\int dE\,
\frac{d\sigma_{\alpha,nl}}{dE}\,,  
\label{50}
\end{eqnarray}
where $\Sigma_{nl}$ takes into account the sum of the different 
sub-shell contributions of the element $\alpha$.

\begin{figure}[t!]
\centering
\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/Fig5.eps}
\caption{Mean emitted energy distribution for ionization by impact of
multicharged ions.}
\label{fig:emittedener}
\end{figure}

Extending the simple stoichiometric model for the mean electron energy
calculation, it results
\begin{equation}
\overline{E}_{M}=\frac{\sum\limits_{\alpha}n_{\alpha}
\overline{E_{\alpha}}}{\sum\limits_{\alpha}n_{\alpha}\sigma_{\alpha}}.
\label{70}
\end{equation}
For impact of H$^{+}$ and He$^{+2}$ on water at 1 MeV/amu, we obtain 
\mbox{$\overline{E}_{\text{H}_{2}\text{O}}=43.7$} and 45 eV, while the 
experimental values 
on liquid water by Pimblott and LaVerne \cite{pimblott2007} were found 
to be 51.5 and 52.2 eV, respectively. Just to stress the importance of 
the projectile charge, for O$^{+8}$ on water, at the same impact energy 
1 MeV/amu, we obtained \mbox{$\overline{E}_{\text{H}_{2}\text{O}}=54.7$} eV, 
which is 25\% larger than proton impact result. It is worth noting that 
there should not be any difference if the calculations were carried in 
first Born approximation. We will came back on this issue later.

\begin{figure}[t!]
\centering
\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/Fig6.eps}
\caption{Mean emitted angle distribution for ionization by impact of
multicharged ions.}
\label{fig:emittedang}
\end{figure}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Ionization of water}

%We now return to the calculation of the mean electron energy of H$^{+}$ 
%and He$^{+2}$ in water, $\overline{E}_{\text{H}_2\text{O}}$. 
Considering that the residual charge of each hydrogen in water is 
$q_{\text{H}}=+0.35$~\cite{rappe1991}, we can imply that the residual charge 
of oxygen is $q_{\text{O}}=-0.7$. Then, we could re-write the formula 
of water using Eq.~(\ref{eq:newstoi}) as H$_{1.3}$O$_{1.17}$ (assuming
$n_{\text{O}}^{\text{CDW}}\simeq 4$ and $n_{\text{H}}^{\text{CDW}}\simeq 1$). 
Recalculating the mean electron energy by the impact of H$^{+}$ and 
He$^{+2}$ on water at 1 MeV/amu, we obtain 
$\overline{E}_{\text{H}_2\text{O}}=46.2$ and 47.6 eV. Our results 
are 10\% apart from the experimental values: 51.5 and 
52.2~eV~\cite{pimblott2007}, respectively, which is not bad considering 
the simplicity of our model.

We can be compute the fraction of energy carried by 
the electron $f$, which is defined as
\begin{equation}
f=\frac{\sum\limits_{\alpha}n_{\alpha}\sigma_{\alpha}
\overline{E_{\alpha}}}{\sum\limits_{\alpha}n_{\alpha}(\sigma_{\alpha}
\overline{E_{\alpha}}+\sum\limits_{nl}\sigma_{\alpha nl}E_{\alpha nl})}\,.
\label{90}
\end{equation}
Using our shell-to-shell ionization cross sections 
$\sigma_{\alpha nl}$~\cite{miraglia2019}, and the atomic binding energies
$E_{\alpha nl}$~\cite{clementi} of the atom $\alpha$, we obtain 
$f=0.70$ and $0.71$ on water at 1 MeV/amu for H$^{+}$ and He$^{+2}$, 
which differ from the respective experimental values $f=0.81$ and $0.80$.

The difference between our calculations and the experimental values 
could be attributed to the binding energy values of the molecular 
water~12.65 eV with respect to its atomic components, H (13.6 eV) and 
O (17.19 eV). This theoretical discrepancy goes beyond our independent 
atom approximation and it cannot be accounted for by our model. Once 
again, we could estimate this contribution by resorting to the 
stoichiometric model. We can make an assesment for a maximum difference 
per molecule of water as 
\begin{equation}
\delta E=(13.6-12.65)\times1.17+(17.19-12.65)\times1.3=7\,\text{eV}  
\label{80}
\end{equation}
Most of this energy should be transfer to the kinetic energy of the 
emitted electron. Thus, by adding 7 eV to the previous values, we obtain
$\overline{E}_{\text{H}_2\text{O}}=53.2$ and 54.6 eV for H$^{+}$ and 
He$^{+2}$ on water at 1 MeV/amu, respectively, which are much closer to
the experimental values. If we add $\delta E$ to the numerator of 
Eq.~(\ref{90}), we obtain $f=0.81$ and $0.80$, in perfect agreement with 
the experiments (see Table 1 of Ref.~\cite{pimblott2007}).

>>>>

\title{Ionization of biological molecules by multicharged ions by using 
the stoichiometric model}
\author{author}
\affil{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio (CONICET-UBA).
Casilla de correo 67, sucursal 28 (C1428EGA) Buenos Aires, Argentina.}
%\author{author}
%\affil{Instituto de Astronom\'{\i}a y F\'{\i}sica del Espacio (CONICET-UBA).
%Casilla de correo 67, sucursal 28 (C1428EGA) Buenos Aires, Argentina.}
%\affil{Fac. de Ciencias Exactas y Naturales, Universidad de Buenos Aires}
\date{\today }

@@

The damage caused by the impact of multicharged heavy projectiles on 
biological targets has become a field of interest due to its recent 
implementation in ion--beam cancer therapy. The effectiveness of the 
radiation depends on the choice of the ions. In particular, theoretical 
and experimental studies with different projectiles have concluded that 
charged carbon ions could be the most suitable ions to be used. The 
study of such systems represent a challenge from the theoretical point 
of view. The most widely used method to compute ionization of 
multicharged atoms is the first Born approximation. This perturbative 
method warrants the $Z^{2}$ laws, where $Z$ is the projectile charge. 
However, the damage is concentrated in the vicinities of the Bragg peak,
at energies of hundreds of keV/amu. Precisely in this intermediate 
region, the Born approximation starts to fail. Another theoretical 
issue arises from the targets themselves; we are dealing with complex 
molecules, which are a hard task for {\it ab initio} calculations. 
The objective of this article is to deal with these two aspects. 
First, we perform more appropriate calculations on the primary damage 
mechanism, i.e., atomic ionization by multicharged ions, which can 
replace the Born results. Second, we inspect and use an stoichiometric 
model, which reduces a molecule to a sum of atomic processes quantities 
weighted by the number of such atom in the molecule.

@@

In a given biological medium, direct ionization by ion impact accounts 
for just a fraction of the overall damage. Secondary electrons, as well 
as recoil target ions, also contribute substantially to the total damage. 
We can consider the single differential cross section of the shell 
$nl$ of the atom $\alpha$, $d\sigma_{\alpha nl}/dE$, to be a function 
of the ejected electron energy $E$ as a simple distribution 
function~\cite{surdutovic2018}. Then, we can define the mean value 
$\overline{E}_{\alpha}$ as in Ref.~\cite{abril2015},
\begin{eqnarray}
\overline{E}_{\alpha} &=&\frac{\langle E_{\alpha}\rangle}{\langle
1\rangle}=\frac{1}{\sigma_{\alpha}}\sum\limits_{nl}\int dE\,E
\frac{d\sigma_{\alpha,nl}}{dE}\,,  
\label{40} \\
\langle 1\rangle &=&\sigma_{\alpha}=\sum\limits_{nl}\int dE\,
\frac{d\sigma_{\alpha,nl}}{dE}\,,  
\label{50}
\end{eqnarray}
where $\Sigma_{nl}$ takes into account the sum of the different 
sub-shell contributions of the element $\alpha$.

\begin{figure}[t!]
\centering
\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/fig5.eps}
\caption{Mean emitted energy distribution for ionization by impact of
multicharged ions. Charge state of projectiles given by arrow.}
\label{fig:emittedener}
\end{figure}

\begin{figure}[t!]
\centering
\includegraphics[width=0.75\textwidth]{figuras/Fig_finales/fig6.eps}
\caption{Mean emitted angle distribution for ionization by impact of
multicharged ions.}
\label{fig:emittedang}
\end{figure}

@@

Lets us consider a molecule $M$ composed by $n_{\alpha}$ atoms of the
element $\alpha$, the SSM describes the total ionization cross section 
of the molecule $\sigma_{M}$ as a simple sum of ionization cross 
sections of the isolated atoms $\sigma_{\alpha}$, 
\begin{equation}
 \sigma_{M}=\sum\limits_{\alpha}n_{\alpha}\sigma_{\alpha}\,.  
 \label{eq:sumion}
\end{equation}
We divided seventeen molecular targets of our interest in three families. 
The classification defined is given in Table~\ref{tab:families}.

@@

The first attempt to develop a comprehensive but straightforward 
phenomenological model for electron ejection from large molecules was 
proposed by Toburen {\it et al.}~\cite{toburen1975,toburen1976}. 
The authors found it convenient to scale the experimental ionization 
cross section in terms of the number of active or weakly--bound valence 
electrons $n_e$ (i.e. total number of electrons minus the K--shell).
Following Toburen, we can define the ionization cross section per weakly 
bound electron, $\sigma_{e}^T$, as
\begin{equation}
\sigma_{e}^T=\frac{\sigma_{M}}{n_e}=\frac{\sum\limits_{\alpha}
n_{\alpha}\sigma_{\alpha}}{\sum\limits_{\alpha}n_{\alpha}\nu_{\alpha}^T}
=\sigma_{e}^T(v)\,, 
\label{27} 
\end{equation}
where $\nu_{\alpha}^T$ are the Toburen numbers for each $\alpha$ atom 
is given by
\begin{equation}
\nu_{\alpha}^T=\left\{ 
\begin{array}{ll}
1, & \text{for H,} \\
4, & \text{for C,} \\ 
5, & \text{for N and P,} \\ 
6, & \text{for O and S}\,.
\end{array}\right.
\label{eq:nelec} 
\end{equation} 
The Toburen rule can be stated by saying that 
$\sigma_{e}$ is a \textit{universal} parameter independent on the 
molecule, which depends solely on the impact velocity, and holds for 
high impact energies (i.e. 0.25--5 Mev/amu).
These $\nu_{\alpha}^T$ can be interpreted as the number of active 
electrons in the collision. Of course, at very high energies also de 
K--shell electrons will be ionized and these numbers will be different.
A similar dependence with the number of weakly bound electrons was 
found in Ref.~\cite{itoh2013} for proton impact on uracil and adenine.

@@

@@

\begin{table}[H]
\begin{center}
\begin{tabular}{llllll}
\hline
 $E_{\text{av}}$ & Adenine & Cytosine & Guanine & Thymine & Uracil \\
\hline
 Molecular & -7.1955 & -6.9058 & -7.4725 & -7.5304 & -7.6224 \\
 Toburen   & -8.4236 & -8.6743 & -8.7275 & -8.7947 & -9.0022 \\
 Our work  & -7.9027 & -7.8639 & -7.9420 & -7.8066 & -7.8839 \\
\hline
\end{tabular}
\caption{.}
\label{tab:gravener}
\end{center}
\end{table}

@@
